The impact of early massive mergers on the chemical evolution of Milky Way-like galaxies: insights from NIHAO-UHD simulations

### 2023-05-11 src/tex/ms.tex
[showyourwork] overleaf sync

Galaxy: structure --- Galaxy: evolution --- Galaxy: kinematics and dynamics --- galaxies:
  formation --- Galaxy: disk --- methods: numerical
 \end{keywords}

@@

\begin{figure*}
    \script{merger_ratios.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/merger_ratios.pdf}
        \vspace*{-1.75em}
        \caption{Gas mass (blue line, left axis) and gas merger ratio (orange line, right axis) as a function of cosmic time. The gray shaded areas mark the times when mergers heavily impact the gas disk size (see also Fig.~\ref{fig:half_mass}). Most of these mergers are at early times (2-4 Gyr), with the exceptions of g7.08e11 and g2.79e12, which also have a late time merger each.}
        \label{fig:merger_ratio}
    \end{centering}
\end{figure*}

@@

For this work we use four simulations from the NIHAO-UHD suite \citep{Buck2020a} part of the Numerical Investigation of a Hundred Astronomical Objects (NIHAO) simulation suite \citep{Wang2015}. Parts of this simulation suite have previously been used to study the build-up of MW's peanut-shaped bulge \citep{Buck2018,Buck2019b}, investigate the stellar bar properties \citep{Hilmi2020}, infer the MW's dark halo spin \citep{Obreja2022}, study the dwarf galaxy inventory of MW mass galaxies \citep{Buck2019} or investigate the age-metallicity relation of MW disk stars \citep{Lu2022} including the chemical bimodality of disk stars \citep{Buck2020}, their abundances \citep{Lu2022a} and the origin of very metal-poor stars inside the stellar disk \citep{Sestito2021}.
Comparing the properties of these galaxies with observations of the MW and local disk galaxies from the SPARC data \citep{Lelli2016}, \citet{Buck2020a} showed that simulated galaxy properties agree well with observations.

@@

\begin{figure}
    \script{gas_profile_all.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/gas_profile_2d.pdf}
        \vspace*{-1.75em}
        \caption{
            Relative change in cold gas surface density profile before and after the merger events as indicated in Fig.~\ref{fig:merger_ratio} by the gray vertical lines. Outside of a radius of $\sim5-10$ kpc the merging satellites contribute significant amounts of cold gas up to 200\%.
        }
        \label{fig:surf_den}
    \end{centering}
\end{figure}

@@

The sudden increase of gas mass due to these gas rich mergers leads to a sudden increase in the half mass radius ($\gsim20$ kpc) of the cold gas disk of the main galaxy as the orange lines (right axis) in Figure~\ref{fig:half_mass} show. On the same figure we also plot the metallicity gradient of the cold gas disk of the main galaxy as a function of time (blue lines, left axis), measured by fitting a straight line to the radial metallicity profile in the radial range $2.5 - 17.5$ kpc. In order to minimize short time scale fluctuations we plot the running average with a window size of 10 corresponding to a time window of $\sim600$~Myr. 
%One may raise question to whether a linear fit is sufficient during merger events, \cite{Lu2022Rblim} has investigated this and show that after the stellar disk has formed, there always exist a strong linear correlation between [Fe/H] and R.

We find that the time evolution of the metallicity gradient in the galaxies shows a rather complex behaviour. Galaxies g2.79e12 and g8.26e11 show a fast flattening of the initial metallicity gradient at early times ($\lesssim2$ Gyr and $\lesssim 2.5$ Gyr) from initially $-0.08$ dex/kpc to $-0.04$ dex/kpc while g7.55e11 and g7.08e11 start out already with a shallow gradient of $\sim-0.05$ and $\sim-0.04$.
Common to all four galaxies we find a steepening of the metallicity gradient by at least $0.02$ dex/kpc over a time frame of $\sim1$ Gyr at times where the gas disk size rapidly increases (marked with gray vertical lines) due to the increasing gas mass from the merging satellites. After the steepening happened, the gradient flattens again by $\sim0.01-0.02$ dex/kpc over roughy the same time frame except for g2.79e12 for which the gradient continues to steepen after the merger up to a maximum of $\sim-0.1$ dex/kpc. Only after another $\sim1.5$ Gyr the gradient starts to quickly flatten again until the present day. The late time evolution of the gradient of all four galaxies is comparatively weak but it consistently continues to flatten with at most a decrease of the gradient of $\sim0.02$ dex/kpc over a time of $\sim6-7$ Gyr.

@@

In order to investigate the cause of the strong steepening of the metallicity gradient in our simulations we first plot in Figure~\ref{fig:surf_den} where in the disk the accreted gas mostly contributes. Figure~\ref{fig:surf_den} plots the relative change of cold gas surface mass density radial profile before the gas rich merger and after the merger finished as indicated by the gray vertical lines in Figure~\ref{fig:half_mass}. This figure reveals an interesting finding. Consistently across all four galaxies the surface mass density of cold gas in the inner $5$ kpc is roughly constant or decreases slightly due to gas consumption by star formation and gas expulsion by feedback. Outside of $5$ kpc the surface mass density increases by $\sim75$ \% up to $\sim200$\% due to cold gas added by the merger event, either by direct accretion of cold gas from the merging satellite or by triggered gas accretion from the circum-galactic medium (CGM) due to tidal forces exerted by the interacting satellite. Note, the increase fraction correlates strongly with radius. Accretion of less enriched gas from the hotter CGM might be due to cold gas coagulation (whose seed can be the cold gas from the merging satellites or fountain gas \citep[e.g.][]{Armillotta2016,Sparre2020} from previous SNe/stellar winds) triggered by (gravitational) perturbations \citep[e.g.][]{Gronke2022}.

@@

Figure~\ref{fig:feh_evolution} complements our findings from Fig.~\ref{fig:half_mass} and shows that during phases of strong steepening of the metallicity gradient (gray vertical lines) it is especially the outskirts of the galaxies (outside of the solar circle, $R>8$ kpc) that lack behind the central parts in [Fe/H] enrichment. In general, we find that the inner annulus of $3<R<5$ kpc (black line) evolves almost self-similar to the central galaxy and relatively unaffected by the mergers. The outer galaxy annuli (blue, orange and red lines) show strong negative deviations/dips at the times of mergers (compare also to Fig.~\ref{fig:merger_ratio}), or similarly at the times of strong increase in the cold gas half-mass radius (see Fig.~\ref{fig:half_mass}), which indicates that the cold [Fe/H]-enriched gas at these radii is strongly influenced by the more metal-poor gas brought in by the merging dwarf galaxies. We note that after the merger, the elevated star formation in the galaxy outskirts causes these radii to quickly catch up in [Fe/H] enrichment \citep[see also Fig. 7 in][]{Buck2020}.

@@

 Figure~\ref{fig:mdf} shows the metallicity distribution function (MDF) for all the gas (orange histogram) and stars (blue histogram) inside the virial radius of four gas-rich satellites that merge with the main galaxy at $2.02$ Gyr, $2.57$ Gyr (two mergers happening at this time) and $9.27$ Gyr in comparison to the MDF of gas (gray histogram) and stars (black histogram) in the central galaxy at the corresponding times. A similar figure for the [$\alpha$/Fe] distribution function is shown in Fig.~\ref{fig:mdf_oxygen} of the appendix. Focusing first on the stellar components (step histograms) we see that the stars contributed by the mergers have on average $\sim0.75$ dex lower metallicity compared to the stars of the main galaxy although at the lowest metallicities ($\lesssim-2.5$) the tails of both distributions overlap. Turning to the gaseous component (filled histograms) we find that (i) the gas and stars of the dwarf galaxies have rather similar metallicity distribution functions where the gas seems to have slightly higher metallicity by about $\sim0.1$ dex. The gaseous MDF of the main galaxy on the other hand shows two peaks separated by about $\sim1$ dex. Those two peaks represent the star forming gas disk at higher metallicities between $-1\lesssim \rm{[Fe/H]}\lesssim0$ and the gas of the CGM at metallicities lower than $-1$. Comparing the gas MDF of the merger to the one of the main galaxy we recover the findings from the stellar component; the merger has an MDF that peaks $\sim0.75$ dex below the MDF of the main galaxy's gas disc and thus broadly overlaps with the CGM gas peak of the main galaxy. It is exactly this difference in metal enrichment in combination with the preferential addition of the accreted gas to the galaxy outskirts that causes the steepenig of the metallicity gradient in the main galaxy's star forming gas by about $0.025$ dex as observed in Fig.~\ref{fig:half_mass} by reducing the metallicity in the galaxy outskirts by $\sim0.25$ dex as Fig.~\ref{fig:feh_evolution} shows. Thus, our expectation is that the GSE merger in the MW must have had a similarly lower metallicity at time of merging compared to the MW's gas disc in order to achieve a steepening of the MW's metallcity gradient of $\sim0.03$ dex as measured by \cite{Lu2022b} and Ratcliffe et al. submitted.

@@

\begin{figure*}
    \script{feh_ofe.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/2.79e12_feh_ofe_grid.pdf}
        \vspace*{-1.75em}
        \caption{
            [$\alpha$/Fe] vs. [Fe/H] plane for the stars in the galaxy g2.79e12 (black histogram in the background). Individual panels show with color where the stars contributed by the merging dwarf galaxies reside in this plane (upper row), and where the gas from the merging dwarf galaxies contributes to star formation after the merger completed (lower panel). \T{For better visibility we chose to display filled 2d-histograms for the stars in the upper panels and open contour lines for the gas in the lower panels.} 
        }
        \label{fig:ofe_feh}
    \end{centering}
\end{figure*}

@@

As stated previously, \cite{Lu2022a} and Ratcliffe et al. submitted find a steepening in the stellar [Fe/H] gradient about $8-9$ Gyr ago. Here, using four hydrodynamical simulations, we find that early massive mergers can cause a steepening of the metallicity gradient, supporting their conclusions that the GSE was responsible. One difference between our work here and their work is that we find the gradient of the gas steepens for two out four galaxies for only $\sim$1 Gyr before beginning to flatten out again, while both \cite{Lu2022a} and Ratcliffe et al. submitted find that the stellar metallicity gradient steepens over a time frame of 2+ Gyr. Such a long phase of steepening is only observed in the galaxy g2.79e12 and g7.08e11, for which then also the phase of flattening is longer compared to the other two galaxies in our sample and last from the time of merger until the present day. Understanding the roll of the relative merger mass size, velocity, and metallicity, as well as observational age and [Fe/H] uncertainties, on the impact of this time frame is beyond the scope of this work. Nevertheless, the conclusions drawn here highlight the effect of mergers on galactic disk metallicity gradients, and clarify that the [Fe/H] gradient in the MW disk does not monotonically flatten with time as assumed in some previous works (e.g. \citealt{Kubryk2015, Minchev2018}).

Fig 6 suggests that the merger history of the MW should leave a unique signature in its AMR. \cite{Ciuca2022} identify a ``dip" and a diagonal feature increasing in [Fe/H] with cosmic time which they call the ``Great Galactic Starburst" in the MW's denoised AMR. Our results support their conclusions that these features were caused by an early massive merger (GSE). They discuss that these features are different regions of the thick disk (high-$\alpha$ sequence). We now understand that this ``starburst" and dip episode are due to the merging of the second branch with the main galaxy's branch after the early merger finished.

In addition to the primary steepening phase just after an early massive merger, we find additional steepening phases in the [Fe/H] gradient in each simulation (Figure 2).  Since a large percentage of the gas is stripped during the first crossing of a merger (\citealt{TepperGarcia2018}), we would expect subsequent passages to have a weaker effect on the gradient. Looking at g2.79e12, which has a late time merger at $\sim$9.3 Gyr, with pericenter passages at $\sim$10 and 12 Gyr, we see that the first pericenter passage has a stronger effect on the metallicity gradient than the second passage (Figure 2). This mimics the findings in observational data by Ratcliffe et al. submitted, who discovered three fluctuations in the metallicity gradient, and attributed the later two fluctuations to Sagittarius’ pericenter passages. If Sagittarius was the culprit, we would expect a second track below the main galaxy's in the AMR if there is a late time merger (Fig. 6). This signature may be visible in some age catalogs (\citealt{Xiang2022}, Anders et al. in prep), however current age and metallicity uncertainties mask the distinctness of the two tracks. Still, our results suggest that Sgr may be responsible for these features both in the AMR and the metallicity gradient. Future work with precise and accurate stellar ages and metallicity could use these multiple steepenings in the metallicity gradient to constrain passage times and potentially the process of gas stripping of merging satellites.

Finally, \cite{Annem2022} find that the effect of gas infall from Sagittarius is primarily felt in the outer disk, where the gas is diluted in [Fe/H]. This agrees with our Fig 4, where we directly show that the dilution in [Fe/H] is felt more in the outer disk, and rarely felt near the galactic center.

@@

>>>>

Galaxy: structure --- Galaxy: evolution --- Galaxy: kinematics and dynamics --- galaxies:  formation --- galaxies: ISM --- methods: numerical
 \end{keywords}

@@

\begin{figure*}
    \script{merger_ratios.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/merger_ratios.pdf}
        \vspace*{-1.75em}
        \caption{Gas mass (blue line, left axis) and gas merger ratio (orange line, right axis) as a function of cosmic time. The gray shaded areas mark the times when mergers heavily impact the gas disk size (see also Fig.~\ref{fig:half_mass}) \textbf{chosen such that they bracket the time when the satellites cross the virial radius of the main galaxy and the time the satellite has fully coalescent with the main galaxy}. Most of these mergers are at early times (2-4 Gyr), with the exceptions of g7.08e11 and g2.79e12, which also have a late time merger each.}
        \label{fig:merger_ratio}
    \end{centering}
\end{figure*}

@@

For this work we use four simulations from the NIHAO-UHD suite \citep{Buck2020a} part of the Numerical Investigation of a Hundred Astronomical Objects (NIHAO) simulation suite \citep{Wang2015}. \textbf{These 4 galaxies are chosen to reflect the most MW like galaxies in terms of mass, size and diks properties.} Parts of this simulation suite have previously been used to study the build-up of MW's peanut-shaped bulge \citep{Buck2018,Buck2019b}, investigate the stellar bar properties \citep{Hilmi2020}, infer the MW's dark halo spin \citep{Obreja2022}, study the dwarf galaxy inventory of MW mass galaxies \citep{Buck2019} or investigate the age-metallicity relation of MW disk stars \citep{Lu2022} including the chemical bimodality of disk stars \citep{Buck2020}, their abundances \citep{Lu2022a} and the origin of very metal-poor stars inside the stellar disk \citep{Sestito2021}.
Comparing the properties of these galaxies with observations of the MW and local disk galaxies from the SPARC data \citep{Lelli2016}, \citet{Buck2020a} showed that simulated galaxy properties agree well with observations.

@@

\begin{figure*}
    \begin{centering}
     \includegraphics[width=\linewidth]{static/frame0019.pdf}
     \includegraphics[width=\linewidth]{static/frame0023.pdf}
     \includegraphics[width=\linewidth]{static/frame0027.pdf}
     \includegraphics[width=\linewidth]{static/frame0038.pdf}
     \includegraphics[width=\linewidth]{static/frame0044.pdf}
        \vspace*{-1.75em}
        \caption{

        }
        \label{fig:half_mass}
    \end{centering}
\end{figure*}

\begin{figure}
    \script{gas_profile_all.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/gas_profile_2d.pdf}
        \vspace*{-1.75em}
        \caption{
            Relative change in cold gas surface density profile before and after the merger events as indicated in Fig.~\ref{fig:merger_ratio} by the gray shaded areas. Outside of a radius of $\sim5-10$ kpc the merging satellites contribute significant amounts of cold gas up to 200\%.
        }
        \label{fig:surf_den}
    \end{centering}
\end{figure}

@@

The sudden increase of gas mass due to these gas rich mergers leads to a sudden increase in the half mass radius ($\gsim20$ kpc) of the cold gas disk of the main galaxy as the orange lines (right axis) in Figure~\ref{fig:half_mass} show. On the same figure we also plot the metallicity gradient of the cold gas disk of the main galaxy as a function of time (blue lines, left axis), measured by fitting a straight line to the radial metallicity profile in the radial range $2.5 - 17.5$ kpc. \textbf{We have chosen this radial range such that it encompasses the star forming disk to highlight the effect of gas accretion on the central disk. Note, this radial range is sometimes smaller than the cold gas $R_{\rm half}$ during times of the merger which can be biased high by a non-spherical gas distribution like bridges etc (see e.g. Fig.~XX).} In order to minimize short time scale fluctuations we plot the running average with a window size of 10 corresponding to a time window of $\sim600$~Myr. 
%One may raise question to whether a linear fit is sufficient during merger events, \cite{Lu2022Rblim} has investigated this and show that after the stellar disk has formed, there always exist a strong linear correlation between [Fe/H] and R.

We find that the time evolution of the metallicity gradient in the galaxies shows a rather complex behaviour. Galaxies g2.79e12 and g8.26e11 show a fast flattening of the initial metallicity gradient at early times ($\lesssim2$ Gyr and $\lesssim 2.5$ Gyr) from initially $-0.08$ dex/kpc to $-0.04$ dex/kpc while g7.55e11 and g7.08e11 start out already with a shallow gradient of $\sim-0.05$ and $\sim-0.04$.
Common to all four galaxies we find a steepening of the metallicity gradient by at least $0.02$ dex/kpc over a time frame of $\sim1$ Gyr at times where the gas disk size rapidly increases (marked with gray shaded areas) due to the increasing gas mass from the merging satellites. After the steepening happened, the gradient flattens again by $\sim0.01-0.02$ dex/kpc over roughy the same time frame except for g2.79e12 for which the gradient continues to steepen after the merger up to a maximum of $\sim-0.1$ dex/kpc. Only after another $\sim1.5$ Gyr the gradient starts to quickly flatten again until the present day. The late time evolution of the gradient of all four galaxies is comparatively weak but it consistently continues to flatten with at most a decrease of the gradient of $\sim0.02$ dex/kpc over a time of $\sim6-7$ Gyr.

@@

In order to investigate the cause of the strong steepening of the metallicity gradient in our simulations we first plot in Figure~\ref{fig:surf_den} where in the disk the accreted gas mostly contributes. Figure~\ref{fig:surf_den} plots the relative change of cold gas surface mass density radial profile before the gas rich merger and after the merger finished as indicated by the \textbf{gray shaded areas} in Figure~\ref{fig:half_mass}. This figure reveals an interesting finding. Consistently across all four galaxies the surface mass density of cold gas in the inner $5$ kpc is roughly constant or decreases slightly due to gas consumption by star formation and gas expulsion by feedback. Outside of $5$ kpc the surface mass density increases by $\sim75$ \% up to $\sim200$\% due to cold gas added by the merger event, either by direct accretion of cold gas from the merging satellite or by triggered gas accretion from the circum-galactic medium (CGM) due to tidal forces exerted by the interacting satellite. Note, the increase fraction correlates strongly with radius. Accretion of less enriched gas from the hotter CGM might be due to cold gas coagulation (whose seed can be the cold gas from the merging satellites or fountain gas \citep[e.g.][]{Armillotta2016,Sparre2020} from previous SNe/stellar winds) triggered by (gravitational) perturbations \citep[e.g.][]{Gronke2022}.

@@

Figure~\ref{fig:feh_evolution} complements our findings from Fig.~\ref{fig:half_mass} and shows that during phases of strong steepening of the metallicity gradient (\textbf{gray shaded areas}) it is especially the outskirts of the galaxies (outside of the solar circle, $R>8$ kpc) that lack behind the central parts in [Fe/H] enrichment. In general, we find that the inner annulus of $3<R<5$ kpc (black line) evolves almost self-similar to the central galaxy and relatively unaffected by the mergers. The outer galaxy annuli (blue, orange and red lines) show strong negative deviations/dips at the times of mergers (compare also to Fig.~\ref{fig:merger_ratio}), or similarly at the times of strong increase in the cold gas half-mass radius (see Fig.~\ref{fig:half_mass}), which indicates that the cold [Fe/H]-enriched gas at these radii is strongly influenced by the more metal-poor gas brought in by the merging dwarf galaxies. We note that after the merger, the elevated star formation in the galaxy outskirts causes these radii to quickly catch up in [Fe/H] enrichment \citep[see also Fig. 7 in][]{Buck2020}.

@@

 Figure~\ref{fig:mdf} shows the metallicity distribution function (MDF) for all the gas (orange histogram) and stars (blue histogram) inside the virial radius of four gas-rich satellites that merge with the main galaxy at $2.02$ Gyr, $2.57$ Gyr (two mergers happening at this time) and $9.27$ Gyr in comparison to the MDF of gas (gray histogram) and stars (black histogram) in the central galaxy at the corresponding times. A similar figure for the [$\alpha$/Fe] distribution function is shown in Fig.~\ref{fig:mdf_oxygen} of the appendix. Focusing first on the stellar components (step histograms) we see that the stars contributed by the mergers have on average $\sim0.75$ dex lower metallicity compared to the stars of the main galaxy although at the lowest metallicities ($\lesssim-2.5$) the tails of both distributions overlap. Turning to the gaseous component (filled histograms) we find that (i) the gas and stars of the dwarf galaxies have rather similar metallicity distribution functions where the gas seems to have slightly higher metallicity by about $\sim0.1$ dex. The gaseous MDF of the main galaxy on the other hand shows two peaks separated by about $\sim1$ dex. Those two peaks represent the star forming gas disk at higher metallicities between $-1\lesssim \rm{[Fe/H]}\lesssim0$ and the gas of the CGM at metallicities lower than $-1$. Comparing the gas MDF of the merger to the one of the main galaxy we recover the findings from the stellar component; the merger has an MDF that peaks $\sim0.75$ dex below the MDF of the main galaxy's gas disc and thus broadly overlaps with the CGM gas peak of the main galaxy. It is exactly this difference in metal enrichment in combination with the preferential addition of the accreted gas to the galaxy outskirts that causes the steepenig of the metallicity gradient in the main galaxy's star forming gas by about $0.025$ dex \textbf{kpc$^{-1}$} as observed in Fig.~\ref{fig:half_mass} by reducing the metallicity in the galaxy outskirts by $\sim0.25$ dex as Fig.~\ref{fig:feh_evolution} shows. Thus, our expectation is that the GSE merger in the MW must have had a similarly lower metallicity at time of merging compared to the MW's gas disc in order to achieve a steepening of the MW's metallcity gradient of $\sim0.03$ dex as measured by \cite{Lu2022b} and Ratcliffe et al. submitted.

@@

\begin{figure*}
    \script{feh_ofe.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/2.79e12_feh_ofe_grid.pdf}
        \vspace*{-1.75em}
        \caption{
            [$\alpha$/Fe] vs. [Fe/H] plane for the stars in the galaxy g2.79e12 (black histogram in the background). Individual panels show with color where the stars contributed by the merging dwarf galaxies reside in this plane (upper row), and where the gas from the merging dwarf galaxies contributes to star formation after the merger completed (lower panel). 
        }
        \label{fig:ofe_feh}
    \end{centering}
\end{figure*}

@@

As stated previously, \cite{Lu2022a}, Ratcliffe et al. submitted and Wang et al. submitted find a steepening in the stellar [Fe/H] gradient about $8-9$ Gyr ago. \textbf{A steep metallicity gradient for such old stars seems surprising as the old, high-$\alpha$ stars in the MW do not show any significant metallicity gradient at present-day. It is now well established that [Fe/H] gradients of mono-age populations flatten from the effect of radial migration. This flattening is most prominent for the oldest stars, as shown, for example, by \citet[][their Fig.5]{Minchev2013}. Thus, while high-alpha stars have a flat [Fe/H] gradient today, there is a large scatter around it, related to the effect of radial migration \citep[see for example][Fig. 1]{Anders2017}. Thus, this steepening phase can only be revealed using the birth radius of stars as any kind of orbit reshuffling, like e.g. radial migration, will wash out the signal at present-day.}

Here, using four hydrodynamical simulations, we find that early massive mergers can cause a steepening of the metallicity gradient, supporting their conclusions that the GSE was responsible. \textbf{GSE is indeed expected to have been a major merger of about 4:1, or 20\% of the MW mass \citep{Helmi2018} and most likely gas-rich, as expected for mergers at high redshift. An attempt to estimate its gas fraction at the time of merging was done by \citep{Vincenzo2019}, finding a value of ~0.67.}

One difference between our work here and the work of Ratcliffe et al. and \cite{Lu2022a} is that we find the gradient of the gas steepens for two out four galaxies for only $\sim$1 Gyr before beginning to flatten out again, while both \cite{Lu2022a} and Ratcliffe et al. submitted find that the stellar metallicity gradient steepens over a time frame of 2+ Gyr. Such a long phase of steepening is only observed in the galaxy g2.79e12 and g7.08e11, for which then also the phase of flattening is longer compared to the other two galaxies in our sample and last from the time of merger until the present day. Understanding the roll of the relative merger mass size, velocity, and metallicity, as well as observational age and [Fe/H] uncertainties, on the impact of this time frame is beyond the scope of this work. Nevertheless, the conclusions drawn here highlight the effect of mergers on galactic disk metallicity gradients, and clarify that the [Fe/H] gradient in the MW disk does not monotonically flatten with time as assumed in some previous works (e.g. \citealt{Kubryk2015, Minchev2018}).

Figure~\ref{fig:age_feh} suggests that the merger history of the MW should leave a unique signature in its AMR. \cite{Ciuca2022} identify a ``dip" and a diagonal feature increasing in [Fe/H] with cosmic time which they call the ``Great Galactic Starburst" in the MW's denoised AMR. Our results support their conclusions that these features were caused by an early massive merger (GSE). They discuss that these features are different regions of the thick disk (high-$\alpha$ sequence). We now understand that this ``starburst" and dip episode are due to the merging of the second branch with the main galaxy's branch after the early merger finished.

In addition to the primary steepening phase just after an early massive merger, we find additional steepening phases in the [Fe/H] gradient in each simulation (Fig.~\ref{fig:half_mass}).  Since a large percentage of the gas is stripped during the first crossing of a merger (\citealt{TepperGarcia2018}), we would expect subsequent passages to have a weaker effect on the gradient. Looking at g2.79e12, which has a late time merger at $\sim$9.3 Gyr, with pericenter passages at $\sim$10 and 12 Gyr, we see that the first pericenter passage has a stronger effect on the metallicity gradient than the second passage (Fig.~\ref{fig:half_mass}). This mimics the findings in observational data by Ratcliffe et al. submitted, who discovered three fluctuations in the metallicity gradient, and attributed the later two fluctuations to Sagittarius’ pericenter passages. If Sagittarius was the culprit, we would expect a second track below the main galaxy's in the AMR if there is a late time merger (Fig.~\ref{fig:age_feh}). This signature may be visible in some age catalogs \citep{Xiang2022,Anders2023}, however current age and metallicity uncertainties mask the distinctness of the two tracks. Still, our results suggest that Sgr may be responsible for these features both in the AMR and the metallicity gradient. Future work with precise and accurate stellar ages and metallicity could use these multiple steepenings in the metallicity gradient to constrain passage times and potentially the process of gas stripping of merging satellites.

Finally, \cite{Annem2022} find that the effect of gas infall from Sagittarius is primarily felt in the outer disk, where the gas is diluted in [Fe/H]. This agrees with our Fig.~\ref{fig:feh_evolution}, where we directly show that the dilution in [Fe/H] is felt more in the outer disk, and rarely felt near the galactic center.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Data Availability}
This manuscript has been typeset using showyourwork! by \citet{Luger2021} and thus is fully reproduceable. All data and code to recreate the analysis presented here can be found at \url{https://github.com/TobiBu/GSE_merger}. The simulations underlying this work are publicly available at \url{https://tobias-buck.de/#sim_data}.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
